Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-15 aistats.tex
Start editing bias-variance

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
\textcolor{red}{The explanation here is not clear enough, or too concise} 
Now we can match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\aligns{
	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
	\conj(\m) = - \half \log(\m)  \; ,
}
up to constants.
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m_*; \m_n)
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
	\bregman( \nat_n; \nat_* )
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in \cref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptote~\eqref{eq:asymptote} .
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
Note that this inequality actually holds for the symmetrized Bregman $\cB(a,b) + \cB(b,a)$, which explains why it loses a factor 2 compared to the asymptote~\eqref{eq:asymptote}, but we leave this to the appendix for clarity.
Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth
This correlates well with our numerical observations (cf App.\ref{app:gaussian-variance}).

@@

These functions are neither smooth, nor strongly convex, nor self-concordant (see \cref{app:gaussian}).
Let us keep this in mind, as we move on to discussing the general problem, and some ways to solve it via direct expansions of the Bregman divergence.

@@

\subsection{Asymptote}
\label{ssec:asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now detail a classical asymptotic result.
Proofs are in \cref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.

@@

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
From Taylor expansion~\eqref{eq:bregmanTaylor},
we know that all Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance\footnote{
In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in \cref{app:self-concordant}.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
}
\end{proposition}
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.

@@

These results partially answer~\eqref{problem}. 
Partially, because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
None of them apply to small $n$.
That is one caveat. Another one is the gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
A full solution to~\eqref{problem} would address both of these caveats.
Informed by the properties we have seen so far, let us investigate a bias-variance decomposition to formulate a conjecture.
%If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.

\subsection{Bias-Variance Decomposition}
\label{ssec:bias-variance}
In both the quadratic~\eqref{eq:MAP_quadratic} and the Gaussiance variance examples~\eqref{eq:MAP_rate}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, 
giving us a flavor of what we would like as a general result for exponential families : a finite sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
In fact, such a decomposition exists for any Bregman divergence.
Let $\tilde \theta_n := \expecti{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be its corresponding mean parameter.
As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes into
\begin{equation}
	\expect{\bregmanconj(\m^* ; \hat \m_n)} = {\bregmanconj(\m^* ; \tilde \m_n)}
	+ {\expect{\bregmanconj(\tilde \m_n ; \MAPm)}}
	\label{eq:primal_pivot}
\end{equation}

\textbf{Remark:} In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point. An estimator will be unbiased if its primal expectation is non-zero.
However, this is not true for the the gaussian variance MLE, where the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.

@@

\paragraph{Illustrations.}
We plot the bias-variance decompositions for $\cN(0,\sigma^2)$ (Figure~\ref{fig:variance_decomposition}) and  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:thales} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) . \rlp{Tune figures.}\textcolor{red}{Not super clear, what is the "characters featured in these decompositions?"}

\begin{figure}[t]
	\centering
	\includegraphics[width=.4\textwidth]{figs/gaussian_variance.pdf}
	\caption{
	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound.
	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
	}
	\label{fig:variance_decomposition}
\end{figure}

\begin{figure}[t]
	\centering
	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
	\caption{
	\textbf{Full gaussian} $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. \textbf{Left:} training curves (5,95) percentiles, average and asymptote.
	\textbf{Right:} bias-variance decomposition.
	}
	\label{fig:gaussian_decomposition}
\end{figure}

@@

>>>>

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$. 
Now we match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\aligns{
	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
	\conj(\m) = - \half \log(\m)  \; ,
}
up to constants.
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m_*; \m_n)
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) \; .
%	\bregman( \nat_n; \nat_* )
%	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in App.\ref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptote~\eqref{eq:asymptote} .
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
Note that this inequality was derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus was more tractable.
This explains the loss of a factor 2 compared to the asymptote~\eqref{eq:asymptote}. 
Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth
This correlates well with our numerical observations (cf App.\ref{app:gaussian-variance}).

@@

These functions are neither smooth, nor strongly convex, nor self-concordant (see App.\ref{app:gaussian}).
Let us keep this in mind, as we move on to discussing the general problem, and some ways to solve it via direct expansions of the Bregman divergence.

@@

\subsection{Asymptote}
\label{ssec:asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now detail a classical asymptotic result.
Proofs are in App.\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.

@@

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
From Taylor expansion~\eqref{eq:bregmanTaylor},
we know that all Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance\footnote{
In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.\ref{app:self-concordant}.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
}
\end{proposition}
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.

@@

These results partially answer~\eqref{problem}. 
Partially, because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
None of them apply to small $n$.
That is one caveat. Another one is the gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
A full solution to~\eqref{problem} would address both of these caveats.
Informed by the properties we have seen so far, let us formulate a conjecture.
%If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.

\subsection{Bias-Variance Decomposition}
\label{ssec:bias-variance}
In both the quadratic~\eqref{eq:MAP_quadratic} and the Gaussiance variance examples~\eqref{eq:MAP_rate}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, 
giving us a flavor of what we would like as a general result for exponential families : a finite sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
In fact, such a decomposition exists for any Bregman divergence \citep[Theorem 0.1]{pfau2013generalized}.
\begin{theorem}[Bregman Bias-Variance Decomposition]
	Let $\tilde \theta_n := \expecti{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be its dual representation. The  expected Bregman decomposes into
\begin{equation}
	\expect{\bregmanconj(\m^* ; \hat \m_n)} = {\bregmanconj(\m^* ; \tilde \m_n)}
	+ {\expect{\bregmanconj(\tilde \m_n ; \MAPm)}}
	\label{eq:primal_pivot}
\end{equation}
\end{theorem}

\textbf{Remark:} In this decomposition, the primal expectation $\expecti{\hat \theta_n}$ is the reference point. 
An estimator will be unbiased if $\tilde \nat_n = \nat^*$.
This is not true for the MLE, which is unbiased w.r.t. the dual parameter $\expecti{\hat \m_n}=\m^*$.

We show in App.\ref{app:bias-variance} that the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$ for Gaussian variance MAP, and
 ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{ \norm{\m^* - \m_0}^2 }{(1 + \frac{n}{n_0})^2 }$ for quadratics.

@@

\paragraph{Illustrations.}
We plot the bias-variance decompositions  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:thales} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) . \rlp{Tune figures.}\textcolor{red}{Not super clear, what is the "characters featured in these decompositions?"}

%\begin{figure}[t]
%	\centering
%	\includegraphics[width=.4\textwidth]{figs/gaussian_variance.pdf}
%	\caption{
%	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound.
%	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
%	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
%	}
%	\label{fig:variance_decomposition}
%\end{figure}

\begin{figure}[t]
	\centering
	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
	\caption{
	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$.
	} %TODO this figure has no labels.t
	\label{fig:gaussian_decomposition}
\end{figure}

@@

\section{BIAS-VARIANCE}
\label{app:bias-variance}
